Update Meeting: 23-09-2021

### 2021-09-24 analysis/update_meeting_23_09_2021.Rmd
publish update meeting 24-09-21

options(scipen=999)
source("code/settings.R")
library(targets)
library(dplyr)
library(DT)
library(knitr)
library(ggplot2)

@@

- Two tasks to explore:
    1. Identify overlap between $X_i \rightarrow X_p$ IVs and $Y_i \rightarrow Y_p$. 
    2. Adjust the $Y_i \rightarrow Y_p$ MR results for effects of $X_i$ (as depicted below).

@@

- Is simply calculating the SNP overlap good enough? 
- Better to calculate signal overlap using LD at some $r^2$ cutoff (what threshold would be useful?).
  - For each IV used to estimate the causal effect $Y_i \rightarrow Y_p$,
  - Test if they are in LD with any IV used to estimate the causal effect $X_i \rightarrow X_p$ at $r^2$ > `0.9`, 
  - Repeat for all IVs for $Y_i$ to calcualte the percentage overlap between the two sets of IVs. 
  - Define percentage overlap as: (# of $Y_i$ IVs are in LD with at least one $X_i$ IV at $r^2$ > `0.9`) / (total number of $Y_i$ IVs used).

The results can be seen below. The `YiXi_IV_exact_overlap` column indicates the percentage of $Y_i$ IVs that are an exact match with $X_i$ IVs (same `rs#`). The `YiXi_IV_sig_overlap` column denotes the percentage overlap of signals, as defined above.

```{r XiYi_IV_overlap, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}
proxyyMR_IV_overlap.DT <- proxyyMR_IV_overlap %>% dplyr::select(exposure_description, outcome_description, YiXi_IV_exact_overlap, YiXi_IV_sig_overlap) %>% unique() %>% arrange(desc(YiXi_IV_sig_overlap))

@@

```{r YiYp_adj, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}
yiyp_adj.DT <- proxyMR_MR_paths_summary_yiyp_adj %>% dplyr::select(exposure_description, outcome_description, exposure_sex, yiyp_IVW_beta, yiyp_IVW_se, yiyp_IVW_pval, yiyp_IVW_beta_adj, yiyp_IVW_se_adj, yiyp_IVW_pval_adj)

@@

fig_dat <- full_join(proxyyMR_IV_overlap, proxyMR_MR_paths_summary_yiyp_adj)

@@

  legend_title <- "IV overlap"
  plot <- ggplot2::ggplot(data = plot_data, ggplot2::aes(x = yiyp_IVW_beta_adj,
                                                         y = yiyp_IVW_beta, color = factor(overlap_above_50))) +
    geom_point(alpha = 3/4) +

    geom_smooth(mapping = aes(x = yiyp_IVW_beta_adj, y = yiyp_IVW_beta), method = "lm", se=FALSE, formula = y~x+0, fullrange=TRUE, color = custom_col[4]) +
    geom_abline(slope=1, intercept=0, color = "black") +
    scale_color_manual(values = custom_col[c(1,2)], labels=c("overlap > 50%","overlap =< 50%")) +
    theme_minimal() +
    ggplot2::labs(colour = legend_title, x = paste0("MR estimate (Y-index to Y-partner)"),
                  y = paste("AM MR estimate adjusted for IV effect on relevant X"))
}

create_fig(fig_dat, custom_col)

@@

## Updated results with adjusted $Y_i$

@@

t1 <- as.data.frame(table(proxyMR_comparison_summary_yiyp_adj$omega_vs_gam_BF_sig_meta))
colnames(t1) <- c("omega_vs_gam_BF_sig_meta", "Frq")

@@

t2 <- as.data.frame(table(proxyMR_comparison_summary_yiyp_adj$omega_vs_rho_BF_sig_meta))
colnames(t2) <- c("omega_vs_rho_BF_sig_meta", "Frq")

@@

t3 <- as.data.frame(table(proxyMR_comparison_summary_yiyp_adj$omega_vs_gam_rho_BF_sig_meta))
colnames(t3) <- c("omega_vs_gam_rho_BF_sig_meta", "Frq")

@@

## MVMR

Run all MVMR models but need to sum up along all $\beta$ from the MV model to create the $\gamma_{mv}$.

MV MR: $X \rightarrow  Z \rightarrow Y$
  - Drop Z's if not significant in MV MR (i.e. $Y \sim X + Z_1 + Z_2 + Z_3...$)
  - Do we drop from the model itself and recompute the beta-coefficients? Or simply not include them in the result sum?

```{r MV_model, echo = FALSE}
library(knitr)
knitr::include_graphics("assets/MV_trait_model_yiyp_adj.png", error = FALSE)
``

>>>>

options(scipen=999)
source("code/settings.R")
library(targets)
library(dplyr)
library(DT)
library(knitr)
library(ggplot2)
library(kableExtra)

@@

- Two things to explore:
    1. Identify overlap between $X_i \rightarrow X_p$ IVs and $Y_i \rightarrow Y_p$. 
    2. Adjust the $Y_i \rightarrow Y_p$ MR results for effects of $X_i$ (as depicted below).

@@

Is simply calculating the SNP overlap good enough? Better to calculate signal overlap using LD at some $r^2$ cutoff (what threshold would be useful?).

  - For each IV used to estimate the causal effect $Y_i \rightarrow Y_p$,
  - Test if they are in LD with any IV used to estimate the causal effect $X_i \rightarrow X_p$ at $r^2$ > `0.9`, 
  - Repeat for all IVs for $Y_i$ to calcualte the percentage overlap between the two sets of IVs. 
  - Define percentage overlap as: (# of $Y_i$ IVs are in LD with at least one $X_i$ IV at $r^2$ > `0.9`) / (total number of $Y_i$ IVs used).

The results can be seen below. The `YiXi_IV_exact_overlap` column indicates the percentage of $Y_i$ IVs that are an exact match with $X_i$ IVs (same `rs#`). The `YiXi_IV_sig_overlap` column denotes the proportion overlap of signals, as defined above.

```{r XiYi_IV_overlap, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}
proxyyMR_IV_overlap.DT <- proxyyMR_IV_overlap %>% dplyr::select(exposure_description, outcome_description, YiXi_IV_exact_overlap, YiXi_IV_sig_overlap) %>% unique() %>% arrange(desc(YiXi_IV_sig_overlap))

@@

```{r YiYp_adj, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE}
fig_dat <- full_join(proxyyMR_IV_overlap, proxyMR_MR_paths_summary_yiyp_adj)
yiyp_adj.DT <- fig_dat %>% dplyr::select(exposure_description, outcome_description, exposure_sex, YiXi_IV_sig_overlap, yiyp_IVW_beta, yiyp_IVW_se, yiyp_IVW_pval, yiyp_IVW_beta_adj, yiyp_IVW_se_adj, yiyp_IVW_pval_adj)

@@

We can compare the $\alpha_{y_{i}\rightarrow y_{p}}$ before and after adjustment, and clearly see a band of estimates centered around $y = 0$, which matches nearly perfectly the instances where $Y$ IVs share more than 50% of $X$ IVs. There are several cases where the adjusted results caused the $abs(\alpha_{y_{i}\rightarrow y_{p}})$ to **increase**.

```{r YiYp_adj_fig_comparison,  echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE}

create_fig_meta <- function(fig_dat, custom_col){

  legend_title <- "IV overlap"
  plot <- ggplot2::ggplot(data = plot_data, ggplot2::aes(x = yiyp_IVW_meta_beta,
                                                         y = yiyp_IVW_meta_beta_adj, color = factor(overlap_above_50))) +
    geom_point(alpha = 3/4) +

    geom_smooth(mapping = aes(x = yiyp_IVW_meta_beta, y = yiyp_IVW_meta_beta_adj), method = "lm", se=FALSE, formula = y~x+0, fullrange=TRUE, color = custom_col[4]) +
    geom_abline(slope=1, intercept=0, color = "black") +
    scale_color_manual(values = custom_col[c(1,2)], labels=c("overlap =< 50%", "overlap > 50%")) +
    theme_minimal() +
    ggplot2::labs(colour = legend_title, x = paste0("MR estimate\n(Y-index to Y-partner)"),
                  y = paste("AM MR estimate adjusted for\nIV effect on relevant X")) + ggtitle("Comparison of adjusted and unadjusted AM estimated\n(meta-analyzed across sexes)") +
    theme(plot.title = element_text(hjust = 0.5))

  return(plot)
}

@@

  legend_title <- "IV overlap"
  plot <- ggplot2::ggplot(data = plot_data, ggplot2::aes(x = yiyp_IVW_beta,
                                                         y = yiyp_IVW_beta_adj, color = factor(overlap_above_50))) +
    geom_point(alpha = 3/4) +

    geom_smooth(mapping = aes(x = yiyp_IVW_beta, y = yiyp_IVW_beta_adj), method = "lm", se=FALSE, formula = y~x+0, fullrange=TRUE, color = custom_col[4]) +
    geom_abline(slope=1, intercept=0, color = "black") +
    scale_color_manual(values = custom_col[c(1,2)], labels=c("overlap =< 50%", "overlap > 50%")) +
    theme_minimal() +
    ggplot2::labs(colour = legend_title, x = paste0("MR estimate\n(Y-index to Y-partner)"),
                  y = paste("AM MR estimate adjusted for\nIV effect on relevant X"), title = "Comparison of adjusted and unadjusted AM estimated \n (sex-specific results)") + 
    theme(plot.title = element_text(hjust = 0.5))

  return(plot)
}

plot_meta <- create_fig_meta(fig_dat, custom_col)
plot <- create_fig(fig_dat, custom_col)

plot_meta
plot

@@

## Updated two trait results with adjusted $Y_i$.

The results below show the significant difference between $\omega$ vs $\rho$, $\gamma$ and their sum. Concerned that most of the significant results involve cases where the adjusted $abs(\alpha_{y_{i}\rightarrow y_{p}})$ is **greater** than the unadjusted case.

@@

yi_adj_increase <- which(abs(proxyMR_MR_paths_summary_yiyp_adj$yiyp_IVW_meta_beta) < abs(proxyMR_MR_paths_summary_yiyp_adj$yiyp_IVW_meta_beta_adj))

proxyMR_inv <- proxyMR_comparison_summary_yiyp_adj[yi_adj_increase, ]

t1 <- as.data.frame(table(proxyMR_comparison_summary_yiyp_adj$omega_vs_gam_BF_sig_meta)/2)
t1.2 <- as.data.frame(table(proxyMR_inv$omega_vs_gam_BF_sig_meta)/2)

colnames(t1) <- c("omega_vs_gam_BF_sig_meta", "Frq")

@@

t2 <- as.data.frame(table(proxyMR_comparison_summary_yiyp_adj$omega_vs_rho_BF_sig_meta)/2)
t2.2 <- as.data.frame(table(proxyMR_inv$omega_vs_rho_BF_sig_meta)/2)
colnames(t2) <- c("omega_vs_rho_BF_sig_meta", "Frq")

@@

t3 <- as.data.frame(table(proxyMR_comparison_summary_yiyp_adj$omega_vs_gam_rho_BF_sig_meta)/2)
t3.2 <- as.data.frame(table(proxyMR_inv$omega_vs_gam_rho_BF_sig_meta)/2)

colnames(t3) <- c("omega_vs_gam_rho_BF_sig_meta", "Frq")
kbl(t3) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

@@

## MVMR.

Have so far run / am running all MVMR models but need to sum up along all $\beta$ from the MV model to create the $\gamma_{mv}$. (Having some computing issues with pruning.)

Should we use the adjusted $Y_i \rightarrow Y_p$ for calculating $\rho$?

MV MR: $X \rightarrow  Z \rightarrow Y$.

  - Drop Z's if not significant in MV MR (i.e. $Y \sim X + Z_1 + Z_2 + Z_3...$)
  - Do we drop from the model itself and recompute the beta-coefficients? Or simply not include them in the result sum?

```{r MV_model, echo = FALSE}

knitr::include_graphics("assets/MV_trait_model.png", error = FALSE)
```
